Classless Java: Tuning Java Interfaces

### 2015-12-06 paper/paper.tex
Reviewd abstract and intro

This paper shows that, with a simple language feature, default methods
and interface inheritance are in fact very expressive. Our proposed
language feature, called \emph{object interfaces}, enables powerful
object-oriented idioms, using multiple inheritance, to be expressed
conveniently in Java. Object interfaces refine conventional Java
interfaces in three different ways. Firstly, object interfaces have
their own object instantiation mechanism, providing an alternative to
class constructors. Secondly, object interfaces support \emph{abstract
  state operations}, providing a way to use multiple inheritance with
state in Java. Finally, object interfaces allow type refinements that
are often tricky to model in conventional class-based
approaches. Interestingly, object interfaces do not require changes to
the runtime, and they also do not introduce any new syntax (apart from
a small annotation): all three features are achieved by reinterpreting
existing Java syntax, and are translated into regular Java code
without loss of type-safety. Since no new syntax is introduced, it
would be incorrect to call object interfaces a language extension or
syntactic sugar. So we use the term \emph{language tuning} to
characterize this kind of language feature. An implementation of
object interfaces using Java annotations and a formalization of the
static and dynamic semantics are presented. Moreover the usefulness of
object interfaces is illustrated through various examples.

>>>>

This paper shows that, with a simple language feature, default methods
and interface inheritance are in fact very expressive. Our proposed
language feature, called \emph{object interfaces}, enables powerful
object-oriented idioms, using multiple inheritance, to be expressed
conveniently in Java. Object interfaces refine conventional Java
interfaces in three different ways. Firstly, object interfaces have
their own object instantiation mechanism, providing an alternative to
class constructors. Secondly, object interfaces support \emph{abstract
  state operations}, providing a way to use multiple inheritance with
state in Java. Finally, object interfaces allow type refinements that
are often tricky to model in conventional class-based
approaches. Interestingly, object interfaces do not require changes to
the runtime, and they also do not introduce any new syntax: all three features are achieved by reinterpreting
existing Java syntax, and are translated into regular Java code
without loss of type-safety. Since no new syntax is introduced, it
would be incorrect to call object interfaces a language extension or
syntactic sugar. So we use the term \emph{language tuning} to
characterize this kind of language feature. An implementation of
object interfaces using Java annotations and a formalization of the
static and dynamic semantics are presented. Moreover the usefulness of
object interfaces is illustrated through various examples.

### 2015-12-06 paper/sections/Introduction.tex
Reviewd abstract and intro

This paper takes a different approach. Rather than trying to get
around the difficulties by changing the language in fundamental ways,
we show that, with a simple language feature, default methods and
interface inheritance are in fact be very expressive. Our proposed
language feature enables powerful object-oriented idioms, using
multiple inheritance. We call the language feature \emph{object
  interfaces}, because such interfaces can be instantiated directly,
without the need for an explict class definition. Moreover, object
interfaces support \emph{abstract state operations}, providing a way
to use multiple inheritance with state in Java. The abstract state
operations include various common utility methods (such as getters and
setters, or clone-like methods). In the presence of subtyping, such
operations often require special care, as their types need to be
refined. Object interfaces provide support for type-refinement and can
automatically produce code that deals with type-refinement
adequately. %With object interfaces, many Java programs can be built
%without using a single class!

@@

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a syntax-directed translation from CJ to itself. In the
resulting CJ code all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-preserving, ensuring that the translation does not
introduce type-errors. 
%CJ's usefulness goes beyond serving as a
%calculus to formalize object interfaces. During the development
%process of CJ, we encountered a bug in the implementation of default
%methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
%bug, ECJ behaves differently from both our formalization and Oracle's 
%Java compiler.

To evaluate the usefulness of object interfaces, we illustrate 3
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application is shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, all applications are implemented 
without defining a single class!

@@

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove that several properties hold for the translation, including a type-preservation.

@@

\item{\bf Language Tuning} We identify the concept of language tuning 
and we describe object interfaces as an example. We also discuss 
how other existing approaches, such as the annotations in project 
Lombok~\cite{lombok}, which can be viewed as language tuning.

>>>>

This paper takes a different approach. Rather than trying to get
around the difficulties by changing the language in fundamental ways,
we show that, with a simple language feature, default methods and
interface inheritance are in fact very expressive. Our proposed
language feature enables powerful object-oriented idioms, using
multiple inheritance. We call the language feature \emph{object
  interfaces}, because such interfaces can be instantiated directly,
without the need for an explict class definition. Moreover, object
interfaces support \emph{abstract state operations}, providing a way
to use multiple inheritance with state in Java. The abstract state
operations include various common utility methods (such as getters and
setters, or clone-like methods). In the presence of subtyping, such
operations often require special care, as their types need to be
refined. Object interfaces provide support for type-refinement and can
automatically produce code that deals with type-refinement
adequately. %With object interfaces, many Java programs can be built
%without using a single class!

@@

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a syntax-directed translation from CJ to itself. In the
resulting CJ code all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-preserving, ensuring that the translation does not
introduce type-errors in client code. 
%CJ's usefulness goes beyond serving as a
%calculus to formalize object interfaces. During the development
%process of CJ, we encountered a bug in the implementation of default
%methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
%bug, ECJ behaves differently from both our formalization and Oracle's 
%Java compiler.

To evaluate the usefulness of object interfaces, we illustrate 3
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, all applications are implemented 
without defining a single class!

@@

\item {\bf ClassLess Java (CJ):} A simple formal calculus that models 
the essential features of Java 8 interfaces with default methods, and 
can be used to formally define the translation of object interfaces. 
We prove several properties of the translation.

@@

\item{\bf Language Tuning} We identify the concept of language tuning 
and we describe object interfaces as an example. We also discuss 
how other existing approaches, such as the annotations in project 
Lombok~\cite{lombok},  can be viewed as language tuning.
